---
title: 
author: 
date: 
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Required packages. Note: Most of the installations are through CRAN and use the formal install.packages("")
##### The package "phangorn" can be installed via Bioconductor.
```{r, eval=FALSE}
BiocManager::install("phangorn")
install.packages("ape")
install.packages("seqinr")
install.packages("ggtree")
```

##### Load libraries
```{r, message=FALSE, warning=FALSE}
library(ape)
library(phangorn)
library(seqinr)
library(ggtree)
library(ggplot2)
```
##### 1. Read in and explore the alignment
```{r}
seqs <- read.FASTA("fish_align.fasta", type="AA")
```
##### Explore the alignment before doing anything else
```{r}
length(seqs)          # number of sequences
names(seqs)           # represent fish species
seqs                  # summary including sequence lengths
```
##### Convert the sequences to a format phangorn uses
```{r}
dat <- as.phyDat(seqs)
```
##### 2. NJ method builds trees from pairwise distances
#####  Create the distance matrix before building the tree
```{r}
dist_matrix <- dist.ml(dat)
```
##### Look at the raw distances
```{r}
print(dist_matrix)
```
##### 3. Build distance-based tree (NJ) 
```{r}
nj_tree <- NJ(dist_matrix)
```
##### Is the tree rooted or unrooted? 
```{r}
is.rooted(nj_tree)
```

##### Plot the tree
```{r}
plot(nj_tree, type="unrooted", main="Unrooted Neighbor-Joining Tree")
```
##### Now we build a midpoint-rooted tree and plot it
```{r}
nj_rooted <- midpoint(nj_tree)
plot(nj_rooted, main="Midpoint-rooted Neighbor-Joining Tree")
```
##### 4. Now we'll build an ML tree
```{r}
fit <- pml(nj_tree, dat)
```

##### Try a few models
```{r}
fit_JTT <- optim.pml(fit, model = "JTT")
fit_WAG <- optim.pml(fit, model = "WAG")
fit_LG  <- optim.pml(fit, model = "LG")
```
##### Compare models using AIC - which fits the data best?
```{r}
cat("JTT AIC:", AIC(fit_JTT), "\n")
cat("WAG AIC:", AIC(fit_WAG), "\n")
cat("LG AIC:",  AIC(fit_LG),  "\n")
```
##### Use the best fitting model going forward
```{r}
fit_optimized <- fit_JTT
```

##### 5. Compare NJ and ML trees visually
##### Do they agree? Where do they differ?
```{r}
par(mfrow=c(1,2))
plot(midpoint(nj_rooted), main="Neighbor-Joining")
plot(midpoint(fit_optimized$tree), main="Maximum Likelihood (JTT)")
par(mfrow=c(1,1))
```

##### 6. Run bootstrapping
```{r, message=FALSE, warning=FALSE, results='hide'}
bs <- bootstrap.pml(fit_optimized, bs=100)
```
##### Plot with bootstrap values
```{r}
bs_tree <- plotBS(midpoint(fit_optimized$tree), bs, 
                  main="ML Tree with Bootstrap Support")
```
##### How many nodes have strong support (>= 70%)?
```{r}
bs_values <- as.numeric(bs_tree$node.label)
cat("Nodes with bootstrap >= 0.70:", sum(bs_values >= 0.70, na.rm=TRUE), "\n")
cat("Nodes with bootstrap < 0.70:", sum(bs_values < 0.70, na.rm=TRUE), "\n")
```
##### 7. Pretty visualization with ggtree
##### Add a scale bar which is important for interpreting branch lengths
```{r}
ggtree(midpoint(bs_tree)) + 
  geom_tiplab(linewidth=3) + 
  geom_nodelab(aes(label=label), linewidth=3, hjust=1.2) +
  geom_treescale() +
  ggtitle("Maximum Likelihood Phylogeny of Fish Sequences (JTT model)")
```
##### Save an image of your tree!
```{r}
ggsave("fish_tree.png", width=8, height=10, dpi=300)
```
